---
title: "Project 2 — Data Tidying and Transformation"
author: "Radzhana Rabdanova"
date: today
format:
  html:
    toc: true
    code-fold: false
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
---

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(knitr)
```

---

# Dataset 1: FAO Healthy Diet Cost

## Data Source

This dataset is provided by the **Food and Agriculture Organization (FAO)**. It tracks the cost of a healthy diet across different countries and regions from 2017 to 2024.

## Data Structure Before Tidying

The raw data is in a **wide format** where each year (Y2017, Y2018, etc.) is a separate column. This makes it difficult to analyze trends over time.

```{r fao-load}
# Import raw FAO data
fao_raw <- read.csv("fao_diet_raw.csv", check.names = FALSE)

# Show dimensions and column names
cat("Dimensions:", nrow(fao_raw), "rows x", ncol(fao_raw), "columns\n")
cat("Column names:\n")
print(names(fao_raw))
```

```{r fao-preview}
# Preview raw wide-format data
head(fao_raw[, 1:6])
```

## Transformation Steps

We use `pivot_longer()` to convert year columns (Y2017–Y2024) into two columns: `Year` and `Cost`. This transforms the data from wide to tidy (long) format.

```{r fao-tidy}
# Reshape from wide to tidy (long) format
fao_tidy <- fao_raw %>%
  select(Area, Item, starts_with("Y20")) %>%        # Keep ID cols + year cols
  pivot_longer(
    cols      = starts_with("Y20"),                  # Pivot all year columns
    names_to  = "Year",                              # New column: year label
    values_to = "Cost"                               # New column: cost value
  ) %>%
  mutate(
    Year = as.numeric(gsub("Y", "", Year)),          # Remove "Y" prefix → numeric
    Cost = as.numeric(as.character(Cost))            # Ensure Cost is numeric
  ) %>%
  drop_na(Cost)                                      # Remove rows with missing Cost

# Preview tidy data
head(fao_tidy)
```

```{r fao-summary}
# Summary: average cost by year
fao_summary <- fao_tidy %>%
  group_by(Year) %>%
  summarize(
    Avg_Cost = round(mean(Cost, na.rm = TRUE), 2),
    .groups  = "drop"
  )

kable(fao_summary, col.names = c("Year", "Average Diet Cost (USD/day)"),
      caption = "Average Global Cost of a Healthy Diet by Year")
```

## Analysis & Visualization

```{r fao-plot}
ggplot(fao_summary, aes(x = Year, y = Avg_Cost)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title   = "Average Global Cost of a Healthy Diet (2017–2024)",
    x       = "Year",
    y       = "Average Cost (USD per day)"
  ) +
  theme_minimal()
```

## Conclusion

The data shows a **steady increase** in the cost of a healthy diet globally, particularly accelerating after 2020. This trend likely reflects global inflationary pressures and supply chain disruptions following the COVID-19 pandemic.

---

# Dataset 2: Gym Members Exercise Tracking

## Data Source

Sourced from **Kaggle's Gym Members Exercise Tracking** dataset. It tracks physical metrics and workout types for gym members, including heart rate measurements across different training styles.

## Data Structure Before Tidying

The data contains **separate columns** for three heart rate metrics: `Max_BPM`, `Avg_BPM`, and `Resting_BPM`. To compare these across workout types, they need to be in a single column.

```{r gym-load}
# Import raw gym data
gym_raw <- read.csv("gym_members_exercise_tracking.csv")

# Show dimensions
cat("Dimensions:", nrow(gym_raw), "rows x", ncol(gym_raw), "columns\n")

# Preview raw data
head(gym_raw)
```

## Transformation Steps

We use `pivot_longer()` to collapse the three BPM columns into two new columns: `BPM_Type` (category) and `BPM_Value` (measurement).

```{r gym-tidy}
# Select relevant columns and reshape to long format
gym_tidy <- gym_raw %>%
  select(Workout_Type, Max_BPM, Avg_BPM, Resting_BPM) %>%   # Keep only needed cols
  pivot_longer(
    cols      = ends_with("BPM"),                             # Pivot all BPM columns
    names_to  = "BPM_Type",                                   # New column: BPM category
    values_to = "BPM_Value"                                   # New column: numeric value
  )

# Preview tidy data
head(gym_tidy, 12)
```

```{r gym-summary}
# Summary table: average BPM by workout type and BPM type
gym_summary <- gym_tidy %>%
  group_by(Workout_Type, BPM_Type) %>%
  summarize(
    Mean_BPM = round(mean(BPM_Value, na.rm = TRUE), 1),
    .groups  = "drop"
  )

kable(gym_summary, col.names = c("Workout Type", "BPM Type", "Mean BPM"),
      caption = "Average Heart Rate by Workout Type and BPM Category")
```

## Analysis & Visualization

```{r gym-plot}
ggplot(gym_tidy, aes(x = Workout_Type, y = BPM_Value, fill = BPM_Type)) +
  geom_boxplot() +
  labs(
    title  = "Heart Rate Distribution by Workout Type",
    x      = "Workout Category",
    y      = "Beats Per Minute",
    fill   = "BPM Type"
  ) +
  theme_minimal()
```

## Conclusion

HIIT and Cardio workouts consistently show **higher Average and Maximum BPM** values compared to Yoga and Strength training, which aligns with their higher intensity levels. Resting BPM remains stable across all workout types, reflecting baseline fitness rather than workout intensity.

---

# Dataset 3: Global CO2 Emissions

## Data Source

This dataset tracks **CO2 emission rates** (metric tons) per country and year, covering the period from 1990 to 2021.

## Data Structure Before Tidying

The raw data stores emission values as **character strings** rather than numbers, which prevents any mathematical operations. Additionally, year values contain mixed types that require cleaning.

```{r co2-load}
# Import raw CO2 data
co2_raw <- read.csv("tidy_format_co2_emission_dataset.csv")

# Show structure
str(co2_raw)

# Preview raw data
head(co2_raw)
```

## Transformation Steps

We convert character columns to numeric types and filter out incomplete records.

```{r co2-tidy}
# Clean and type-convert the CO2 dataset
co2_clean <- co2_raw %>%
  rename(country   = Country,                              # Consistent lowercase names
         year      = Year,
         emissions = CO2EmissionRate..mt.) %>%
  mutate(
    year      = as.numeric(as.character(year)),            # Convert year to numeric
    emissions = as.numeric(as.character(emissions))        # Convert emissions to numeric
  ) %>%
  filter(!is.na(emissions), !is.na(year))                  # Remove incomplete rows

# Preview clean data
head(co2_clean)
cat("Rows after cleaning:", nrow(co2_clean), "\n")
```

```{r co2-summary}
# Summarize total global emissions per year
co2_summary <- co2_clean %>%
  group_by(year) %>%
  summarize(
    Total_Emissions = round(sum(emissions, na.rm = TRUE), 2),
    .groups = "drop"
  )

kable(co2_summary, col.names = c("Year", "Total Emissions (mt CO2)"),
      caption = "Total Global CO2 Emissions by Year")
```

## Analysis & Visualization

```{r co2-plot}
ggplot(co2_summary, aes(x = year, y = Total_Emissions)) +
  geom_col(fill = "darkred") +
  labs(
    title = "Total Global CO2 Emissions by Year",
    x     = "Year",
    y     = "Total mt CO2"
  ) +
  theme_minimal()
```

## Conclusion

Global CO2 emissions remained **relatively stable** between approximately 900–1,050 metric tons throughout the 1990s and 2000s. A noticeable dip around **2009** corresponds to reduced industrial activity during the global financial crisis. The sharp spike in 2021 likely reflects expanded country coverage in that year's data collection rather than a literal emissions explosion — an important data quality consideration.

---

# Overall Conclusions

By applying tidy data principles — specifically reshaping wide-format datasets into long format using `pivot_longer()` — we created reproducible, analysis-ready datasets across three distinct domains:

- **FAO Diet Data**: revealed rising global food costs post-2020
- **Gym Exercise Data**: confirmed heart rate patterns align with workout intensity
- **CO2 Emissions Data**: illustrated long-term emissions trends with a data quality observation for 2021

This workflow demonstrates that clean, tidy data is not just aesthetically preferable — it **unlocks the analysis**.
